---
title: "Splicing Analysis"
author: "Daniel Guevara"
date: "2025-05-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Call libraries
```{r}
library(scales)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(ggplot2)
library(ggrepel)
library(patchwork)
library(pheatmap)
library(openxlsx)
library(OmicsKit)
```

# 2. Setup and sample metadata
```{r}
setwd("~/Documents/SimonLab/HLA-I_Analysis/splicing")
library_size_df <- read.csv("library_size.tsv", sep = "\t")

# Create metadata
sampledata <- data.frame(id = c("RU51_N6", "RU57_N1x", "RU59_N7", "RU63_N2", "RU148_N",
                                "RU51_P4", "RU51_M7", "RU51_M16", "RU51_M49", "RU57_M1x", "RU59_P5", "RU59_M6",
                                "RU62_M1x", "RU63_R2", "RU88_R8", "RU88_M15", "RU117_M10", "RU123_MT", "RU148_T8", "RU148_T11"),
                         patient_internal = c(51, 57, 59, 63, 148,
                                              rep(51, 4), 57, rep(59, 2),
                                              62, 63, rep(88, 2), 117, 123, rep(148, 2)),
                         patient = c(1:3, 5, 9,
                                     rep(1, 4), 2, rep(3, 2),
                                     4, 5, rep(6, 2), 7, 8, rep(9, 2)),
                         tumor = c(rep("N", 5), rep("T", 15)),
                         tumor_type = c(rep("N", 5), "P", rep("M", 4), "P",
                                        rep("M", 2), "R", rep("M", 4), rep("P", 2))) %>%
  
  mutate(patient_internal = factor(patient_internal, levels = c(51, 57, 59, 62, 63, 88, 117, 123, 148)),
         patient = factor(patient, levels = 1:9), tumor = factor(tumor, levels = c("N", "T")),
         tumor_type = factor(tumor_type, levels = c("N", "P", "R", "M")),
         internal_tumor = paste0("RU", patient_internal, " ", tumor),
         internal_type = paste0("RU", patient_internal, " ", tumor_type),
         label_tumor = paste0("R0", patient, " ", tumor), label_type = paste0("R0", patient, " ", tumor_type))

# Manage repeated labels
sampledata[, 6:9] <- lapply(sampledata[, 6:9], function(x) {
  ave(x, x, FUN = function(z) {
    if (length(z) == 1) return(z)
    paste0(z, seq_along(z))
  })
})

# Add library size
sampledata <- left_join(sampledata, library_size_df, by = "id")
```

# 3. Overview of splicing events
## 3.1 Pie charts and stacked barplots
```{r}
my_colors <- c("#F4F7AE", "#E6BDF4", "#FC9272", "#8E90D8", "#B9F1D6")

summary_all <- read.csv("results/variable/output/summary.txt", header = TRUE, sep = "\t")
summary_all_nov <- read.csv("results/variable_novel/output/summary.txt", header = TRUE, sep = "\t")

# Stacked bar plots
summary_all <- summary_all %>% transmute(dataset = "Canonical", event = EventTypeDescription,
                                         Total = TotalEventsJCEC, Significant = SignificantEventsJCEC)

summary_all_nov <- summary_all_nov %>% transmute(dataset = "Novel", event = EventTypeDescription,
                                                 Total = TotalEventsJCEC, Significant = SignificantEventsJCEC)

summary_long <- bind_rows(summary_all, summary_all_nov) %>%
  pivot_longer(c(Total, Significant), names_to = "metric", values_to = "n") %>%
  mutate(dataset = factor(dataset, levels = c("Canonical", "Novel")),
         metric = factor(metric, levels = c("Total", "Significant"),
                         labels = c("Total events", "Significant (FDR < 0.05)")),
         event = factor(event, levels = c("skipped exon", "mutually exclusive exons", "retained intron",
                                          "alternative 3' splice sites", "alternative 5' splice sites")))

ggplot(summary_long, aes(x = dataset, y = n, fill = event)) +
  geom_col(color = "black", linewidth = 0.1) +  theme_bw() +
  facet_wrap(~ metric, nrow = 1, scales = "free_y") +
  labs(x = NULL, y = "Frecuency", fill = "Splicing event") +
  scale_y_continuous(labels = scales::comma, limits = c(0, NA), expand = expansion(mult = c(0, 0.01))) +
  scale_fill_manual(values = my_colors, labels = c("Skipped Exon", "Mutually Exclusive Exons", "Retained Intron",
                                                   "Alt. 3' Splice Site", "Alt. 5' Splice Site")) +
  theme(strip.background = element_blank(),
        strip.text = element_text(size = 16),
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.y = element_text(size = 16),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12))


# Pie charts
#source("src/nice_pie.R")

# Canonical splicing events
#p.pie.tot <- nice_pie(summary_all, selected_cols = c("EventType", "TotalEventsJCEC"), colors = my_colors) +
#  theme(legend.position = "none")
#p.pie.sig <- nice_pie(summary_all, selected_cols = c("EventType", "SignificantEventsJCEC"), colors = my_colors)

#p.pie <- p.pie.tot + p.pie.sig

# Novel splicing events
#p.pie.tot.nov <- nice_pie(summary_all_nov, selected_cols = c("EventType", "TotalEventsJCEC"), colors = my_colors) +
#  theme(legend.position = "none")
#p.pie.sig.nov <- nice_pie(summary_all_nov, selected_cols = c("EventType", "SignificantEventsJCEC"), colors = my_colors)

#p.pie.nov <- p.pie.tot.nov + p.pie.sig.nov
```

## 3.2. Correlation of count models [JC vs JCEC]
```{r}
#source("src/psi_correlation.R")

#psi_correlation(df1 = res.SE.JC[res.SE.JC$ID %in% res.SE$ID, ],
#                df2 = res.SE.JCEC[res.SE.JCEC$ID %in% res.SE$ID, ], title = "Skipped Exon")
#psi_correlation(df1 = res.MXE.JC[res.MXE.JC$ID %in% res.MXE$ID, ],
#                df2 = res.MXE.JCEC[res.MXE.JCEC$ID %in% res.MXE$ID, ], title = "Mutually Exclusive Exons")
#psi_correlation(df1 = res.RI.JC[res.RI.JC$ID %in% res.RI$ID, ],
#                df2 = res.RI.JCEC[res.RI.JCEC$ID %in% res.RI$ID, ], title = "Retained Intron")
#psi_correlation(df1 = res.A3SS.JC[res.A3SS.JC$ID %in% res.A3SS$ID, ],
#                df2 = res.A3SS.JCEC[res.A3SS.JCEC$ID %in% res.A3SS$ID, ], title = "Alt. 3' Splice Site")
#psi_correlation(df1 = res.A5SS.JC[res.A5SS.JC$ID %in% res.A5SS$ID, ],
#                df2 = res.A5SS.JCEC[res.A5SS.JCEC$ID %in% res.A5SS$ID, ], title = "Alt. 5' Splice Site")

#ggsave("figures/Correlation_JC_JCEC_Skipped_Exon.jpg", width = 5.5, height = 5, dpi = 300)
#ggsave("figures/Correlation_JC_JCEC_Mutually_Exclusive_Exons.jpg", width = 5.5, height = 5, dpi = 300)
#ggsave("figures/Correlation_JC_JCEC_Retained_Intron.jpg", width = 5.5, height = 5, dpi = 300)
#ggsave("figures/Correlation_JC_JCEC_Alt_3_Splice_Site.jpg", width = 5.5, height = 5, dpi = 300)
#ggsave("figures/Correlation_JC_JCEC_Alt_5_Splice_Site.jpg", width = 5.5, height = 5, dpi = 300)
```

# 4. Managing results
## 4.1. Read data
```{r}
# Annotated splicing events
#res.SE <- read.csv("results/variable/output/SE.MATS.JCEC.txt", sep = "\t")
#res.MXE <- read.csv("results/variable/output/MXE.MATS.JCEC.txt", sep = "\t")
#res.RI <- read.csv("results/variable/output/RI.MATS.JCEC.txt", sep = "\t")
#res.A3SS <- read.csv("results/variable/output/A3SS.MATS.JCEC.txt", sep = "\t")
#res.A5SS <- read.csv("results/variable/output/A5SS.MATS.JCEC.txt", sep = "\t")

# Get list of IDs
canon_SE <- read.csv("results/variable/output/SE.MATS.JCEC.txt", sep = "\t")$ID
canon_MXE <- read.csv("results/variable/output/MXE.MATS.JCEC.txt", sep = "\t")$ID
canon_RI <- read.csv("results/variable/output/RI.MATS.JCEC.txt", sep = "\t")$ID
canon_A3SS <- read.csv("results/variable/output/A3SS.MATS.JCEC.txt", sep = "\t")$ID
canon_A5SS <- read.csv("results/variable/output/A5SS.MATS.JCEC.txt", sep = "\t")$ID

# Novel splicing events
res.SE.nov <- read.csv("results/variable_novel/output/SE.MATS.JCEC.txt", sep = "\t")
res.MXE.nov <- read.csv("results/variable_novel/output/MXE.MATS.JCEC.txt", sep = "\t")
res.RI.nov <- read.csv("results/variable_novel/output/RI.MATS.JCEC.txt", sep = "\t")
res.A3SS.nov <- read.csv("results/variable_novel/output/A3SS.MATS.JCEC.txt", sep = "\t")
res.A5SS.nov <- read.csv("results/variable_novel/output/A5SS.MATS.JCEC.txt", sep = "\t")
```

## 4.2. Counts distribution and filtering
```{r}
# Get Counts Per Million (Inclusion Counts for Retained Intron events)
source("src/get_skipping_counts.R")

SE.CPM <- get_skipping_counts(res.SE.nov, sampledata)
MXE.CPM <- get_skipping_counts(res.MXE.nov, sampledata)
RI.CPM <- get_skipping_counts(res.RI.nov, sampledata, preffix = "IJC")
A3SS.CPM <- get_skipping_counts(res.A3SS.nov, sampledata)
A5SS.CPM <- get_skipping_counts(res.A5SS.nov, sampledata)

# Plot distributions
source("src/nice_Hist.R")
hist_thresholds <- c(seq(0.05, 0.25, 0.05), 0.4, 0.5)

nice_Hist(SE.CPM, thresholds = hist_thresholds, xlim = 25) + labs(x = "CPM - Skipping Counts", title = "Skipped Exon")
nice_Hist(MXE.CPM, thresholds = hist_thresholds, xlim = 25) + labs(x = "CPM - Skipping Counts", title = "Mutually Exclusive Exons")
nice_Hist(RI.CPM, thresholds = hist_thresholds, xlim = 25) + labs(x = "CPM - Inclusion Counts", title = "Retained Intron")
nice_Hist(A3SS.CPM, thresholds = hist_thresholds, xlim = 25) + labs(x = "CPM - Skipping Counts", title = "Alt. 3' Splice Site")
nice_Hist(A5SS.CPM, thresholds = hist_thresholds, xlim = 25) + labs(x = "CPM - Skipping Counts", title = "Alt. 5' Splice Site")

combined_hist <- rbind(SE.CPM[,23:26], MXE.CPM[,23:26], RI.CPM[,23:26], A3SS.CPM[,23:26], A5SS.CPM[,23:26])
combined_hist$event <- c(rep("Skipped Exon", nrow(SE.CPM)),
                         rep("Mutually Exclusive Exons", nrow(MXE.CPM)),
                         rep("Retained Intron", nrow(RI.CPM)),
                         rep("Alt. 3' Splice Site", nrow(A3SS.CPM)),
                         rep("Alt. 5' Splice Site", nrow(A5SS.CPM)))
combined_hist$event <- factor(combined_hist$event, levels = unique(combined_hist$event), ordered = T)

nice_Hist(counts = combined_hist, thresholds = hist_thresholds, xlim = 25) +
  facet_wrap(~event, scales = "free_y", nrow = 1) + xlab ("Counts Per Million") +
  theme(strip.background = element_blank(), strip.text = element_text(size = 14))

ggsave("figures/CPM_Distribution_per_Event.jpg", width = 22, height = 5, dpi = 600)

cpm_threshold <- 0.25
cts_threshold <- cpm_threshold*mean(sampledata$lib_size)/(10**6)

# Filter low count events
source("src/mean_preprocessing.R")

## IJC + SJC
#res.SE.jc <- mean_preprocessing(res.SE, c("IJC_SAMPLE_1", "SJC_SAMPLE_1", "IJC_SAMPLE_2", "SJC_SAMPLE_2"), 65)
#res.MXE.jc <- mean_preprocessing(res.MXE, c("IJC_SAMPLE_1", "SJC_SAMPLE_1", "IJC_SAMPLE_2", "SJC_SAMPLE_2"), 65)
#res.RI.jc <- mean_preprocessing(res.RI, c("IJC_SAMPLE_1", "SJC_SAMPLE_1", "IJC_SAMPLE_2", "SJC_SAMPLE_2"), 65)
#res.A3SS.jc <- mean_preprocessing(res.A3SS, c("IJC_SAMPLE_1", "SJC_SAMPLE_1", "IJC_SAMPLE_2", "SJC_SAMPLE_2"), 65)
#res.A5SS.jc <- mean_preprocessing(res.A5SS, c("IJC_SAMPLE_1", "SJC_SAMPLE_1", "IJC_SAMPLE_2", "SJC_SAMPLE_2"), 65)

res.SE.nov.jc <- mean_preprocessing(res.SE.nov, c("IJC_SAMPLE_1", "SJC_SAMPLE_1", "IJC_SAMPLE_2", "SJC_SAMPLE_2"), cts_threshold)
res.MXE.nov.jc <- mean_preprocessing(res.MXE.nov, c("IJC_SAMPLE_1", "SJC_SAMPLE_1", "IJC_SAMPLE_2", "SJC_SAMPLE_2"), cts_threshold)
res.RI.nov.jc <- mean_preprocessing(res.RI.nov, c("IJC_SAMPLE_1", "SJC_SAMPLE_1", "IJC_SAMPLE_2", "SJC_SAMPLE_2"), cts_threshold)
res.A3SS.nov.jc <- mean_preprocessing(res.A3SS.nov, c("IJC_SAMPLE_1", "SJC_SAMPLE_1", "IJC_SAMPLE_2", "SJC_SAMPLE_2"), cts_threshold)
res.A5SS.nov.jc <- mean_preprocessing(res.A5SS.nov, c("IJC_SAMPLE_1", "SJC_SAMPLE_1", "IJC_SAMPLE_2", "SJC_SAMPLE_2"), cts_threshold)
```

# 3. Managing splicing results [JCEC model]
```{r}
# Sample type PSI
#res.SE.psi <- mean_preprocessing(res.SE.sig, c("IncLevel1", "IncLevel2"), 0.05)
#res.MXE.psi <- mean_preprocessing(res.MXE.sig, c("IncLevel1", "IncLevel2"), 0.05)
#res.RI.psi <- mean_preprocessing(res.RI.sig, c("IncLevel1", "IncLevel2"), 0.05)
#res.A3SS.psi <- mean_preprocessing(res.A3SS.sig, c("IncLevel1", "IncLevel2"), 0.05)
#res.A5SS.psi <- mean_preprocessing(res.A5SS.sig, c("IncLevel1", "IncLevel2"), 0.05)

res.SE.nov.psi <- mean_preprocessing(res.SE.nov.jc, c("IncLevel1", "IncLevel2"), 0.05)
res.MXE.nov.psi <- mean_preprocessing(res.MXE.nov.jc, c("IncLevel1", "IncLevel2"), 0.05)
res.RI.nov.psi <- mean_preprocessing(res.RI.nov.jc, c("IncLevel1", "IncLevel2"), 0.05)
res.A3SS.nov.psi <- mean_preprocessing(res.A3SS.nov.jc, c("IncLevel1", "IncLevel2"), 0.05)
res.A5SS.nov.psi <- mean_preprocessing(res.A5SS.nov.jc, c("IncLevel1", "IncLevel2"), 0.05)

# Filtering FDR and Delta(PSI)
#res.SE.sig <- res.SE.jc %>% filter(FDR < 0.01 & abs(IncLevelDifference) >= 0.05)
#res.MXE.sig <- res.MXE.jc %>% filter(FDR < 0.01 & abs(IncLevelDifference) >= 0.05)
#res.RI.sig <- res.RI.jc %>% filter(FDR < 0.01 & abs(IncLevelDifference) >= 0.05)
#res.A3SS.sig <- res.A3SS.jc %>% filter(FDR < 0.01 & abs(IncLevelDifference) >= 0.05)
#res.A5SS.sig <- res.A5SS.jc %>% filter(FDR < 0.01 & abs(IncLevelDifference) >= 0.05)

res.SE.nov.sig <- res.SE.nov.psi %>% filter(FDR < 0.01 & abs(IncLevelDifference) >= 0.05)
res.MXE.nov.sig <- res.MXE.nov.psi %>% filter(FDR < 0.01 & abs(IncLevelDifference) >= 0.05)
res.RI.nov.sig <- res.RI.nov.psi %>% filter(FDR < 0.01 & abs(IncLevelDifference) >= 0.05)
res.A3SS.nov.sig <- res.A3SS.nov.psi %>% filter(FDR < 0.01 & abs(IncLevelDifference) >= 0.05)
res.A5SS.nov.sig <- res.A5SS.nov.psi %>% filter(FDR < 0.01 & abs(IncLevelDifference) >= 0.05)

# Final data frames
res.SE.nov.sig$event <- "SE"
res.MXE.nov.sig$event <- "MXE"
res.RI.nov.sig$event <- "RI"
res.A3SS.nov.sig$event <- "A3SS"
res.A5SS.nov.sig$event <- "A5SS"

master_SE <- rbind(res.SE.nov.sig[, c("ID", "GeneID", "geneSymbol", "SJC_SAMPLE_1", "SJC_SAMPLE_2",
                                      "IJC_SAMPLE_1", "IJC_SAMPLE_2", "FDR", "IncLevelDifference", "event")],
                   res.MXE.nov.sig[, c("ID", "GeneID", "geneSymbol", "SJC_SAMPLE_1", "SJC_SAMPLE_2",
                                       "IJC_SAMPLE_1", "IJC_SAMPLE_2", "FDR", "IncLevelDifference", "event")],
                   res.RI.nov.sig[, c("ID", "GeneID", "geneSymbol", "SJC_SAMPLE_1", "SJC_SAMPLE_2",
                                      "IJC_SAMPLE_1", "IJC_SAMPLE_2", "FDR", "IncLevelDifference", "event")],
                   res.A3SS.nov.sig[, c("ID", "GeneID", "geneSymbol", "SJC_SAMPLE_1", "SJC_SAMPLE_2",
                                        "IJC_SAMPLE_1", "IJC_SAMPLE_2", "FDR", "IncLevelDifference", "event")],
                   res.A5SS.nov.sig[, c("ID", "GeneID", "geneSymbol", "SJC_SAMPLE_1", "SJC_SAMPLE_2",
                                        "IJC_SAMPLE_1", "IJC_SAMPLE_2", "FDR", "IncLevelDifference", "event")])
table(master_SE$event)
sum(table(master_SE$event))

table(master_SE$event[master_SE$IncLevelDifference < 0])
sum(table(master_SE$event[master_SE$IncLevelDifference < 0]))
```

## 3.3 Stacked bar plots of tumor-specific events
```{r}
source("src/get_recurrent_events.R")

# Subsets of CPMs
SE.CPM.sig <- SE.CPM %>% filter(ID %in% res.SE.nov.sig$ID[res.SE.nov.sig$IncLevelDifference < 0]) %>% mutate(event = "SE")
MXE.CPM.sig <- MXE.CPM %>% filter(ID %in% res.MXE.nov.sig$ID[res.MXE.nov.sig$IncLevelDifference < 0]) %>% mutate(event = "MXE")
RI.CPM.sig <- RI.CPM %>% filter(ID %in% res.RI.nov.sig$ID[res.RI.nov.sig$IncLevelDifference < 0]) %>% mutate(event = "RI")
A3SS.CPM.sig <- A3SS.CPM %>% filter(ID %in% res.A3SS.nov.sig$ID[res.A3SS.nov.sig$IncLevelDifference < 0]) %>% mutate(event = "A3SS")
A5SS.CPM.sig <- A5SS.CPM %>% filter(ID %in% res.A5SS.nov.sig$ID[res.A5SS.nov.sig$IncLevelDifference < 0]) %>% mutate(event = "A5SS")

sig_events <- rbind(SE.CPM.sig, MXE.CPM.sig, RI.CPM.sig, A3SS.CPM.sig, A5SS.CPM.sig) |>
  mutate(event = factor(event, levels = c("SE","MXE","RI","A3SS","A5SS"))) %>% mutate(source = "filtered")

SE.CPM.rec <- get_recurrent_events(SE.CPM, res.SE.nov, sampledata, "SE")
MXE.CPM.rec <- get_recurrent_events(MXE.CPM, res.MXE.nov, sampledata, "MXE")
RI.CPM.rec <- get_recurrent_events(RI.CPM, res.RI.nov, sampledata, "RI")
A3SS.CPM.rec <- get_recurrent_events(A3SS.CPM, res.A3SS.nov, sampledata, "A3SS")
A5SS.CPM.rec <- get_recurrent_events(A5SS.CPM, res.A5SS.nov, sampledata, "A5SS")

rec_events <- rbind(SE.CPM.rec, MXE.CPM.rec, RI.CPM.rec, A3SS.CPM.rec, A5SS.CPM.rec) |>
  mutate(event = factor(event, levels = c("SE","MXE","RI","A3SS","A5SS"))) %>% mutate(source = "recurrent")

all_events <- rbind(sig_events, rec_events) %>%
  mutate(source = factor(source, levels = c("filtered", "recurrent")))
all_events <- all_events[order(all_events$source, decreasing = F), ]
all_events <- all_events[order(all_events$event, decreasing = F), ]

res.SE.tumor <- res.SE.nov %>% filter(ID %in% all_events$ID[all_events$event == "SE"])
res.SE.tumor <- res.SE.tumor[, c(1:11, 17:20, 23)]
res.SE.tumor <- left_join(res.SE.tumor,
                          all_events[all_events$event == "SE", c("ID", "event", "source")], by = "ID")
res.SE.tumor <- res.SE.tumor %>%
  mutate(novel = ifelse(ID %in% canon_SE, "canonical", "novel")) %>% arrange(novel, source)

res.MXE.tumor <- res.MXE.nov %>% filter(ID %in% all_events$ID[all_events$event == "MXE"])
res.MXE.tumor <- res.MXE.tumor[, c(1:13, 19:22, 25)]
res.MXE.tumor <- left_join(res.MXE.tumor,
                           all_events[all_events$event == "MXE", c("ID", "event", "source")], by = "ID")
res.MXE.tumor <- res.MXE.tumor %>%
  mutate(novel = ifelse(ID %in% canon_MXE, "canonical", "novel")) %>% arrange(novel, source)

res.RI.tumor <- res.RI.nov %>% filter(ID %in% all_events$ID[all_events$event == "RI"])
res.RI.tumor <- res.RI.tumor[, c(1:11, 17:20, 23)]
res.RI.tumor <- left_join(res.RI.tumor,
                          all_events[all_events$event == "RI", c("ID", "event", "source")], by = "ID")
res.RI.tumor <- res.RI.tumor %>%
  mutate(novel = ifelse(ID %in% canon_RI, "canonical", "novel")) %>% arrange(novel, source)

res.A3SS.tumor <- res.A3SS.nov %>% filter(ID %in% all_events$ID[all_events$event == "A3SS"])
res.A3SS.tumor <- res.A3SS.tumor[, c(1:11, 17:20, 23)]
res.A3SS.tumor <- left_join(res.A3SS.tumor,
                            all_events[all_events$event == "A3SS", c("ID", "event", "source")], by = "ID")
res.A3SS.tumor <- res.A3SS.tumor %>%
  mutate(novel = ifelse(ID %in% canon_A3SS, "canonical", "novel")) %>% arrange(novel, source)

res.A5SS.tumor <- res.A5SS.nov %>% filter(ID %in% all_events$ID[all_events$event == "A5SS"])
res.A5SS.tumor <- res.A5SS.tumor[, c(1:11, 17:20, 23)]
res.A5SS.tumor <- left_join(res.A5SS.tumor,
                            all_events[all_events$event == "A5SS", c("ID", "event", "source")], by = "ID")
res.A5SS.tumor <- res.A5SS.tumor %>%
  mutate(novel = ifelse(ID %in% canon_A5SS, "canonical", "novel")) %>% arrange(novel, source)

saveRDS(res.SE.tumor, file = "results/SE_tumor_specific.rds")
saveRDS(res.MXE.tumor, file = "results/MXE_tumor_specific.rds")
saveRDS(res.RI.tumor, file = "results/RI_tumor_specific.rds")
saveRDS(res.A3SS.tumor, file = "results/A3SS_tumor_specific.rds")
saveRDS(res.A5SS.tumor, file = "results/A5SS_tumor_specific.rds")
```

[REMAKE THE FOLLOWING PLOTS] 08/20/2025

## CPM barplots of shared events per samples
```{r}
rec_events_plot <- rec_events[rec_events$Mean.T > 10, ]

sample_cols <- names(all_events)[startsWith(names(all_events), "RU")]

agg <- rec_events_plot %>%
  select(geneSymbol, event, all_of(sample_cols)) %>%
  pivot_longer(cols = all_of(sample_cols),
               names_to = "sample",
               values_to = "CPM") %>%
  mutate(CPM = suppressWarnings(as.numeric(CPM))) %>%
  group_by(geneSymbol, event, sample) %>%
  summarise(
    mean_CPM = if (all(is.na(CPM))) NA_real_ else mean(CPM, na.rm = TRUE),
    .groups = "drop"
  )

source("src/plot_gene_bars.R")

out_dir <- "plots/recurrent_gene_bars"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

genes <- unique(rec_events$geneSymbol)

for (g in genes) {
  p <- plot_gene_bars(agg, g, free_y = TRUE)  # set FALSE if you want shared y across facets
  if (!is.null(p)) {
    ggsave(file.path(out_dir, paste0("bars_", g, ".png")), p, width = 12, height = 4, dpi = 300)
  }
}
```


## Genes with shared event
```{r}
events <- list(
  SE = unique(SE.CPM.sig$geneSymbol[SE.CPM.sig$Mean.T > 5]),
  MXE = unique(MXE.CPM.sig$geneSymbol[MXE.CPM.sig$Mean.T > 5]),
  RI = unique(RI.CPM.sig$geneSymbol[RI.CPM.sig$Mean.T > 5]),
  A3SS = unique(A3SS.CPM.sig$geneSymbol[A3SS.CPM.sig$Mean.T > 5]),
  A5SS = unique(A5SS.CPM.sig$geneSymbol[A5SS.CPM.sig$Mean.T > 5])
)

common_genes <- do.call(rbind, lapply(names(events), function(ev) {
  data.frame(geneSymbol = events[[ev]], event = ev, stringsAsFactors = FALSE)
}))

# Data frame for stacked bar plots
all_events <- rbind(SE.CPM.sig[SE.CPM.sig$geneSymbol %in% common_genes$geneSymbol[common_genes$event == "SE"], ],
                    MXE.CPM.sig[MXE.CPM.sig$geneSymbol %in% common_genes$geneSymbol[common_genes$event == "MXE"], ],
                    RI.CPM.sig[RI.CPM.sig$geneSymbol %in% common_genes$geneSymbol[common_genes$event == "RI"], ],
                    A3SS.CPM.sig[A3SS.CPM.sig$geneSymbol %in% common_genes$geneSymbol[common_genes$event == "A3SS"], ],
                    A5SS.CPM.sig[A5SS.CPM.sig$geneSymbol %in% common_genes$geneSymbol[common_genes$event == "A5SS"], ]) |>
  mutate(event = factor(event, levels = c("SE","MXE","RI","A3SS","A5SS")))

cpm_long <- all_events |>
  select(geneSymbol, event, all_of(sample_cols)) |>
  pivot_longer(cols = all_of(sample_cols),
               names_to = "sample", values_to = "CPM")

# Plot
## Per event
event_gene <- cpm_long |>
  group_by(event, geneSymbol) |>
  summarise(mean_CPM = mean(CPM, na.rm = TRUE), .groups = "drop")

ggplot(event_gene, aes(x = geneSymbol, y = mean_CPM, fill = event)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.1) +
  scale_fill_manual(values  = my_colors) +
  labs(x = "Gene", y = "Counts Per Million", fill = "Event") +
  theme_bw() +
  theme(axis.title = element_text(size = 16),
        axis.text.y = element_text(size = 14),
        axis.text.x = element_text(size = 14, angle = 45, hjust = 1),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12))

## Per sample
# Sum CPM per sample × event
event_sample <- cpm_long %>%
  group_by(sample, event) %>%
  summarise(total_CPM = sum(CPM, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    # keep your preferred sample order
    sample = factor(sample, levels = sample_cols),
    event  = factor(event, levels = c("SE","MXE","RI","A3SS","A5SS"))
  )

# Stacked bars: x = sample, fill = event, y = total CPM across genes
ggplot(event_sample, aes(x = sample, y = total_CPM, fill = event)) +
  geom_col(color = "black", linewidth = 0.1) +
  scale_fill_manual(values = my_colors) +
  labs(x = "Sample", y = "Total Counts Per Million", fill = "Event") +
  theme_bw() +
  theme(axis.title = element_text(size = 16),
        axis.text.y = element_text(size = 14),
        axis.text.x = element_text(size = 14, angle = 45, hjust = 1),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12))
```

## 3.1 Distribution of PSI
```{r}
delta_df <- rbind(res.SE.nov[, c("ID", "GeneID", "geneSymbol", "IncLevelDifference", "FDR")],
                  res.MXE.nov[, c("ID", "GeneID", "geneSymbol", "IncLevelDifference", "FDR")],
                  res.RI.nov[, c("ID", "GeneID", "geneSymbol", "IncLevelDifference", "FDR")],
                  res.A3SS.nov[, c("ID", "GeneID", "geneSymbol", "IncLevelDifference", "FDR")],
                  res.A5SS.nov[, c("ID", "GeneID", "geneSymbol", "IncLevelDifference", "FDR")])
delta_df$event <- c(rep("Skipped Exon", nrow(res.SE.nov)),
                    rep("Mutually Exclusive Exons", nrow(res.MXE.nov)),
                    rep("Retained Intron", nrow(res.RI.nov)),
                    rep("Alt. 3' Splice Site", nrow(res.A3SS.nov)),
                    rep("Alt. 5' Splice Site", nrow(res.A5SS.nov)))
delta_df$event <- factor(delta_df$event, levels = unique(delta_df$event))

delta_df <- delta_df %>%
  mutate(sig = -log10(FDR)) %>%
  mutate(sig = if_else(is.infinite(sig), max(sig[is.finite(sig)]) + 1, sig))

sig_df <- delta_df[delta_df$FDR < 0.01, ]

# Statistical analysis
kruskal.test(IncLevelDifference ~ event, data = delta_df)
pairwise.wilcox.test(delta_df$IncLevelDifference, delta_df$event, p.adjust.method = "BH")
FSA::dunnTest(IncLevelDifference ~ event, data = delta_df, method = "bh")

# Plot
max_val <- max(delta_df$sig)

ggplot(sig_df[abs(sig_df$IncLevelDifference) >= 0.05, ], aes(x = event, y = IncLevelDifference)) + theme_bw() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_boxplot(width = 0.5, fill = "grey90", outlier.shape = NA) +
  geom_errorbar(stat = "boxplot", width = 0.7, linewidth = 0.7, color = "blue",
                aes(ymin = after_stat(middle), ymax = after_stat(middle))) +
  geom_jitter(aes(fill = sig), shape = 21, width = 0.15, size = 2.5, alpha = 1, show.legend = TRUE) +
  scale_fill_gradientn(name = expression(-log[10]~"(FDR)"), colors = c("white", "red", "firebrick4"),
                       values = rescale(c(0, max_val-1, max_val)), breaks = seq(0, 12, 3),
                       guide = guide_colorbar(direction = "horizontal", position = "inside",
                                              barwidth = unit(4, "cm"), barheight = unit(0.4, "cm"))) +
  scale_y_continuous(breaks = seq(-0.8,0.8,0.4), limits = c(-0.8, 0.8)) +
  labs(y = "ΔPSI (Tumor – Normal)") +
  theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_blank(), axis.title.y = element_text(size = 14),
        legend.title = element_text(size = 12), legend.text = element_text(size = 10),
        legend.position.inside = c(0.8, 0.9),
        legend.background = element_rect(color = "black"))

ggsave("figures/Delta_PSI_Distribution_Splicing_Events_FDR_0.01.jpg", width = 9, height = 6, dpi = 300)
ggsave("figures/Delta_PSI_Distribution_Splicing_Events_FDR_0.01_PSI_0.05.jpg", width = 9, height = 6, dpi = 300)

ggsave("figures/Delta_PSI_Distribution_Novel_Splicing_Events_FDR_0.01.jpg", width = 9, height = 6, dpi = 300)
ggsave("figures/Delta_PSI_Distribution_Novel_Splicing_Events_FDR_0.01_PSI_0.05.jpg", width = 9, height = 6, dpi = 300)

```

# 5. Volcano plots
```{r}
cutoff_alpha <- 0.05
cutoff_psi <- 0.1

# Plotting function
volcano_plot <- function(results, x_max = 1, y_max = 6, cutoff_signif = 0.05, cutoff_change = 0.1,
                         signif_var, change_var, label_var, title, colors = c("red", "grey70", "blue"))
{
  # Fixed variables
  sig_nice <- cutoff_signif/5
  chg_nice <- cutoff_change*4
  
  # Manage data
  d.volcano <- data.frame(results)
  d.volcano <- d.volcano[!is.na(d.volcano[, signif_var]) & d.volcano[, signif_var] < 1, ]
  d.volcano[d.volcano[, signif_var] < 10**-y_max, signif_var] <- 10**-y_max
  
  d.volcano$colors <- rep("other", nrow(d.volcano))
  d.volcano$colors[(d.volcano[, change_var] >= cutoff_change & d.volcano[, signif_var] < cutoff_signif)] <- "over"
  d.volcano$colors[(d.volcano[, change_var] <= -cutoff_change & d.volcano[, signif_var] < cutoff_signif)] <- "under"
  
  d.volcano$shapes <- rep("nohits", nrow(d.volcano))
  d.volcano$shapes[d.volcano[, signif_var] <= 10**-y_max] <- "hits"
  
  # Condition for labeling
  cond <- d.volcano[, signif_var] < sig_nice & abs(d.volcano[, change_var]) > chg_nice
  
  # Plot
  p.volcano <- ggplot() + theme_bw() +
    scale_y_continuous(expand = c(0.03, 0),
                       limits = c(0, y_max),
                       breaks = seq(0, y_max, by = 2),
                       minor_breaks = seq(0, y_max, by = 1)) +
    scale_x_continuous(expand = c(0.03, 0.03),
                       limits = c(-x_max, x_max),
                       breaks = seq(-x_max, x_max, by = x_max*0.5),
                       minor_breaks = seq(-x_max, x_max, by = x_max*0.1)) +
    
    # Vertical lines and labels:
    geom_vline(xintercept = c(-cutoff_change, cutoff_change), color = "grey20", alpha = 0.8, linetype = 2, linewidth = 1.2) +
    
    # Horizontal line and label:
    geom_hline(yintercept = -log10(cutoff_signif), color = "grey20", alpha = 0.8, linetype = 2, linewidth = 1.2) +
    geom_text(aes(x = -0.8, y = -log10(cutoff_signif) + 0.15, label = paste("q =", cutoff_signif)), color = "grey20", size = 6.2) +
    
    # Datapoints, in different colors:
    geom_point(data = d.volcano,
               aes(x = .data[[change_var]], y = -log10(.data[[signif_var]]), fill = colors, shape = shapes),
               size = 5.5, color = "gray10", alpha = 0.4, show.legend = TRUE) +
    
    # Color settings
    scale_fill_manual(name = "Splicing change",
                      values = c("over" = colors[1], "other" = colors[2], "under" = colors[3]),
                      breaks = c("over", "other", "under"),
                      labels = c("High in Tumor", "No difference", "High in Normal"),
                      guide = guide_legend(override.aes = aes(shape = 21, size = 3, alpha = 1))) +
    
    # Change outliers shapes
    scale_shape_manual(name = "Shape",
                       values = c("hits" = 24, "nohits" = 21),
                       guide = guide_legend(override.aes = list(fill = "white", size = 3, alpha = 1)),
                       labels = c(bquote(FDR <= 10^-.(y_max)), bquote(FDR > 10^-.(y_max)))) +
    
    # Labels for detectable genes
    geom_label_repel(data = d.volcano[cond, ],
                     aes(x = .data[[change_var]], y = -log10(.data[[signif_var]]), label = .data[[label_var]]),
                     inherit.aes = FALSE, parse = FALSE, max.iter = 5000, color = "black", size = 5,
                     segment.alpha = 1.5, box.padding = unit(0.8, "lines"), min.segment.length = unit(0.01, "lines")) +
    
    # Axis labels:
    labs(x = expression("Δ(PSI)"),
         y = expression("-log"[10]*"(FDR)"),
         title = title) +
    
    # Labels size
    theme(plot.title = element_text(size = 26, hjust = 0.5),
          axis.text.x = element_text(size = 22),
          axis.text.y = element_text(size = 22),
          axis.title.x = element_text(size = 26),
          axis.title.y = element_text(size = 26),
          legend.text = element_text(size = 18),
          legend.title = element_text(size= 20, face = "bold"))
  
  return(p.volcano)
}

# JC model volcanos
pv.SE.JC <- volcano_plot(results = res.SE.JC, signif_var = "FDR", change_var = "IncLevelDifference", label_var = "geneSymbol", title = "Skipped Exon")

pv.MXE.JC <- volcano_plot(results = res.MXE.JC, signif_var = "FDR", change_var = "IncLevelDifference", label_var = "geneSymbol", title = "Mutually Exclusive Exons")

pv.RI.JC <- volcano_plot(results = res.RI.JC, signif_var = "FDR", change_var = "IncLevelDifference", label_var = "geneSymbol", title = "Retained Intron")

pv.A3SS.JC <- volcano_plot(results = res.A3SS.JC, signif_var = "FDR", change_var = "IncLevelDifference", label_var = "geneSymbol", title = "Alt. 3' Splice Sites")

pv.A5SS.JC <- volcano_plot(results = res.A5SS.JC, signif_var = "FDR", change_var = "IncLevelDifference", label_var = "geneSymbol", title = "Alt. 5' Splice Sites")

# JCEC model volcanos
pv.SE.JCEC <- volcano_plot(results = res.SE.JCEC, signif_var = "FDR", change_var = "IncLevelDifference", label_var = "geneSymbol", title = "Skipped Exon")

pv.MXE.JCEC <- volcano_plot(results = res.MXE.JCEC, signif_var = "FDR", change_var = "IncLevelDifference", label_var = "geneSymbol", title = "Mutually Exclusive Exons")

pv.RI.JCEC <- volcano_plot(results = res.RI.JCEC, signif_var = "FDR", change_var = "IncLevelDifference", label_var = "geneSymbol", title = "Retained Intron")

pv.A3SS.JCEC <- volcano_plot(results = res.A3SS.JCEC, signif_var = "FDR", change_var = "IncLevelDifference", label_var = "geneSymbol", title = "Alt. 3' Splice Sites")

pv.A5SS.JCEC <- volcano_plot(results = res.A5SS.JCEC, signif_var = "FDR", change_var = "IncLevelDifference", label_var = "geneSymbol", title = "Alt. 5' Splice Sites")
```

# 6. Heatmaps by splicing event
```{r}
ann_colors <- list(Group = c(Normal = "#bbffff", Tumor = "#ffb6c1"))

# Annotation
group_labels <- c(rep("Normal", 5), rep("Tumor", 15))
anot <- data.frame(Group = factor(group_labels))

# Extract psi values
split_psi <- function(df, gene_col = "geneSymbol", psi_cols = c("IncLevel1", "IncLevel2"), num = 10) {
  initial_df <- df[order(df$FDR), ]
  initial_df <- initial_df[1:num, ]
  psi_split <- lapply(initial_df[, psi_cols], function(x) t(sapply(strsplit(x, ","), as.numeric)))
  psi_mat <- do.call(cbind, psi_split)
  rownames(psi_mat) <- initial_df[[gene_col]]
  return(as.data.frame(psi_mat))
}

# Test data frame
mx.SE.JC <- split_psi(res.SE.JC.sig)
mx.SE.JC <- mx.SE.JC[rowSums(is.na(mx.SE.JC)) != ncol(mx.SE.JC), ]
mx.SE.JCEC <- split_psi(res.SE.JCEC.sig)
mx.SE.JCEC <- mx.SE.JCEC[rowSums(is.na(mx.SE.JCEC)) != ncol(mx.SE.JCEC), ]

mx.MXE.JC <- split_psi(res.MXE.JC.sig)
mx.MXE.JC <- mx.MXE.JC[rowSums(is.na(mx.MXE.JC)) != ncol(mx.MXE.JC), ]
mx.MXE.JCEC <- split_psi(res.MXE.JCEC.sig)
mx.MXE.JCEC <- mx.MXE.JCEC[rowSums(is.na(mx.MXE.JCEC)) != ncol(mx.MXE.JCEC), ]

mx.RI.JC <- split_psi(res.RI.JC.sig)
mx.RI.JC <- mx.RI.JC[rowSums(is.na(mx.RI.JC)) != ncol(mx.RI.JC), ]
mx.RI.JCEC <- split_psi(res.RI.JCEC.sig)
mx.RI.JCEC <- mx.RI.JCEC[rowSums(is.na(mx.RI.JCEC)) != ncol(mx.RI.JCEC), ]

mx.A3SS.JC <- split_psi(res.A3SS.JC.sig)
mx.A3SS.JC <- mx.A3SS.JC[rowSums(is.na(mx.A3SS.JC)) != ncol(mx.A3SS.JC), ]
mx.A3SS.JCEC <- split_psi(res.A3SS.JCEC.sig)
mx.A3SS.JCEC <- mx.A3SS.JCEC[rowSums(is.na(mx.A3SS.JCEC)) != ncol(mx.A3SS.JCEC), ]

mx.A5SS.JC <- split_psi(res.A5SS.JC.sig)
mx.A5SS.JC <- mx.A5SS.JC[rowSums(is.na(mx.A5SS.JC)) != ncol(mx.A5SS.JC), ]
mx.A5SS.JCEC <- split_psi(res.A5SS.JCEC.sig)
mx.A5SS.JCEC <- mx.A5SS.JCEC[rowSums(is.na(mx.A5SS.JCEC)) != ncol(mx.A5SS.JCEC), ]

mx.JC.merged <- rbind(mx.SE.JC, mx.MXE.JC, mx.RI.JC, mx.A3SS.JC, mx.A5SS.JC)
mx.JCEC.merged <- rbind(mx.SE.JCEC, mx.MXE.JCEC, mx.RI.JCEC, mx.A3SS.JCEC, mx.A5SS.JCEC)
colnames(mx.JC.merged) <- colnames$label
colnames(mx.JCEC.merged) <- colnames$label

# Plot heatmap
plot_heatmap <- function(df, annotation = anot, annotation_cols = ann_colors) {
  psi_scaled <- t(scale(t(df)))
  rownames(annotation) <- colnames(psi_scaled)
  
  p.heat <- pheatmap(psi_scaled,
                     cluster_cols = FALSE,
                     cluster_rows = FALSE,
                     show_rownames = TRUE, 
                     scale = "none",
                     gaps_row = c(10, 20, 30, 40),
                     gaps_col = c(5),
                     annotation_col = annotation,
                     annotation_colors = annotation_cols,
                     fontsize_row = 8,
                     show_colnames = TRUE)
  
  return(p.heat)
}

ph.JC.merged <- plot_heatmap(mx.JC.merged)
ph.JCEC.merged <- plot_heatmap(mx.JCEC.merged)
```

# 7. [IN PROGRESS] Barplots per gene with splicing events
```{r}
count_shared_events <- function(df, psi_col, group) {
  # Split and convert PSI strings to numeric vectors
  psi_matrix <- t(sapply(strsplit(df[[psi_col]], ","), as.numeric))
  psi_df <- as.data.frame(psi_matrix)
  
  # Assign consistent sample names
  colnames(psi_df) <- paste(group, seq_len(ncol(psi_df)), sep = "_")
  
  # Bind gene symbol and event ID (optional)
  psi_df$geneSymbol <- df$geneSymbol
  psi_df$eventID <- seq_len(nrow(df))  # to track unique splicing events if needed
  
  # Count number of non-NA & non-zero PSI values per event (row)
  count_df <- psi_df[, 1:(ncol(psi_df) - 2)]  # exclude geneSymbol and eventID
  non_zero_counts <- rowSums(!is.na(count_df) & count_df > 0)
  
  # Create frequency table
  frequency_table <- table(non_zero_counts)
  
  return(list(table = frequency_table, matrix = psi_df))
}


```
