---
title: "Splicing Analysis"
author: "Daniel Guevara"
date: "2025-05-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Call libraries
```{r}
library(ggplot2)
library(ggrepel)
library(patchwork)
library(pheatmap)
library(openxlsx)
```

# 2. Setup directories
```{r}
working_dir <- "~/github/HLA-I_Analysis/splicing"
setwd(working_dir)
```

# 3. Retrieve splicing results
```{r}
# JC model
res.SE.JC <- data.frame(read.csv("results/all_samples_variable_ind/output/SE.MATS.JC.txt", sep = "\t"))
res.SE.JC.sig <- res.SE.JC[res.SE.JC$FDR < 0.01 & abs(res.SE.JC$IncLevelDifference) >= 0.05, ]

res.MXE.JC <- data.frame(read.csv("results/all_samples_variable_ind/output/MXE.MATS.JC.txt", sep = "\t"))
res.MXE.JC.sig <- res.MXE.JC[res.MXE.JC$FDR < 0.01 & abs(res.MXE.JC$IncLevelDifference) >= 0.05, ]

res.RI.JC <- data.frame(read.csv("results/all_samples_variable_ind/output/RI.MATS.JC.txt", sep = "\t"))
res.RI.JC.sig <- res.RI.JC[res.RI.JC$FDR < 0.01 & abs(res.RI.JC$IncLevelDifference) >= 0.05, ]

res.A3SS.JC <- data.frame(read.csv("results/all_samples_variable_ind/output/A3SS.MATS.JC.txt", sep = "\t"))
res.A3SS.JC.sig <- res.A3SS.JC[res.A3SS.JC$FDR < 0.01 & abs(res.A3SS.JC$IncLevelDifference) >= 0.05, ]

res.A5SS.JC <- data.frame(read.csv("results/all_samples_variable_ind/output/A5SS.MATS.JC.txt", sep = "\t"))
res.A5SS.JC.sig <- res.A5SS.JC[res.A5SS.JC$FDR < 0.01 & abs(res.A5SS.JC$IncLevelDifference) >= 0.05, ]

# JCEC model
res.SE.JCEC <- data.frame(read.csv("results/all_samples_variable_ind/output/SE.MATS.JCEC.txt", sep = "\t"))
res.SE.JCEC.sig <- res.SE.JCEC[res.SE.JCEC$FDR < 0.01 & abs(res.SE.JCEC$IncLevelDifference) >= 0.05, ]

res.MXE.JCEC <- data.frame(read.csv("results/all_samples_variable_ind/output/MXE.MATS.JCEC.txt", sep = "\t"))
res.MXE.JCEC.sig <- res.MXE.JCEC[res.MXE.JCEC$FDR < 0.01 & abs(res.MXE.JCEC$IncLevelDifference) >= 0.05, ]

res.RI.JCEC <- data.frame(read.csv("results/all_samples_variable_ind/output/RI.MATS.JCEC.txt", sep = "\t"))
res.RI.JCEC.sig <- res.RI.JCEC[res.RI.JCEC$FDR < 0.01 & abs(res.RI.JCEC$IncLevelDifference) >= 0.05, ]

res.A3SS.JCEC <- data.frame(read.csv("results/all_samples_variable_ind/output/A3SS.MATS.JCEC.txt", sep = "\t"))
res.A3SS.JCEC.sig <- res.A3SS.JCEC[res.A3SS.JCEC$FDR < 0.01 & abs(res.A3SS.JCEC$IncLevelDifference) >= 0.05, ]

res.A5SS.JCEC <- data.frame(read.csv("results/all_samples_variable_ind/output/A5SS.MATS.JCEC.txt", sep = "\t"))
res.A5SS.JCEC.sig <- res.A5SS.JCEC[res.A5SS.JCEC$FDR < 0.01 & abs(res.A5SS.JCEC$IncLevelDifference) >= 0.05, ]
```

## Correlation of results
```{r}
psi_correlation <- function(df1, df2, columns = c("ID", "geneSymbol", "IncLevelDifference"),
                            cutoff = 0.4, title, p_size = 2.5, p_shape = 20)
{
  # Integrate data frames
  merged_df <- merge(df1[, columns], df2[, columns],
                     by = columns[1:2], suffixes = c("_JC", "_JCEC"))
  
  # Create label data frame
  label_df <- subset(merged_df,
                     abs(IncLevelDifference_JC) >= cutoff | abs(IncLevelDifference_JCEC) >= cutoff)
  
  label_df$nudge_y <- ifelse(label_df$IncLevelDifference_JCEC > label_df$IncLevelDifference_JC,
                             0.08, -0.08)
  
  # Pearson correlation
  cor_test <- cor.test(merged_df$IncLevelDifference_JC,
                       merged_df$IncLevelDifference_JCEC, method = "pearson")
  r_squared <- round(cor_test$estimate^2, 3)
  
  # Plotting
  plot <- ggplot(merged_df, aes(x = IncLevelDifference_JC, y = IncLevelDifference_JCEC)) +
    geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
    geom_point(alpha = 0.7, size = p_size, shape = p_shape) + theme_bw() +
    annotate("text", x = 0.6, y = -0.6, label = paste0("R² = ", r_squared), size = 4, hjust = 0) +
    geom_label_repel(data = label_df, aes(label = geneSymbol), size = 3,
                     nudge_y = label_df$nudge_y, segment.alpha = 0.5, segment.size = 0.3,
                     box.padding = unit(0.4, "lines"), point.padding = unit(0.3, "lines"),
                     max.overlaps = Inf, show.legend = FALSE) +
    labs(title = title,
         x = expression(Delta*"PSI"*" - Junction Count"),
         y = expression(Delta*"PSI"*" - Junction & Exon Count")) +
    coord_cartesian(xlim = c(min(merged_df$IncLevelDifference_JC) - 0.1,
                             max(merged_df$IncLevelDifference_JC) + 0.1),
                    ylim = c(min(merged_df$IncLevelDifference_JCEC) - 0.1,
                             max(merged_df$IncLevelDifference_JCEC) + 0.1))
  
  return(plot)
}

psi_correlation(df1 = res.SE.JC.sig, df2 = res.SE.JCEC.sig, title = "Skipped Exon")
psi_correlation(df1 = res.MXE.JC.sig, df2 = res.MXE.JCEC.sig, title = "Mutually Exclusive Exons", cutoff = 0.3)
psi_correlation(df1 = res.RI.JC.sig, df2 = res.RI.JCEC.sig, title = "Retained Intron", cutoff = 0.3)
psi_correlation(df1 = res.A3SS.JC.sig, df2 = res.A3SS.JCEC.sig, title = "Alt. 3' Splice Sites", cutoff = 0.2)
psi_correlation(df1 = res.A5SS.JC.sig, df2 = res.A5SS.JCEC.sig, title = "Alt. 5' Splice Sites", cutoff = 0.2)


```

# 4. Overview of splicing events across samples
```{r}
nice_pie <- function(df, selected_cols = c("typeCol", "freqCol"), colors, order = FALSE, levels = NULL) {
  # Manage data
  pie.df <- df[, selected_cols]
  pie.df$percent <- round(pie.df[, 2]*100/sum(pie.df[, 2]), 1)
  pie.df$label <- paste0(pie.df[, 2], "\n", "(", pie.df$percent, "%)")
  
  if (order) {
    pie.df[, 1] <- factor(pie.df[, 1], levels = levels)
    pie.df <- pie.df[order(pie.df[, 1]), ]
    #pie.df$label <- paste0(pie.df[, 1], "\n", pie.df[, 2], "\n", "(", pie.df$percent, "%)")
  }
  
  # Plot
  p.pie <- ggplot(pie.df, aes(x = "", y = .data[[selected_cols[2]]], fill = .data[[selected_cols[1]]])) +
    geom_bar(stat = "identity", width = 1, color = "white") +
    coord_polar("y", start = 0) +
    theme_void() +
    geom_text(aes(label = label),
              position = position_stack(vjust = 0.5),
              size = 4,
              color = "black") +
    scale_fill_manual(values = colors,
                      labels = c("Alt. 3' Splice Sites (A3SS)", "Alt. 5' Splice Sites (A5SS)",
                                 "Mutually Exclusive Exons (MXE)", "Retained Intron (RI)", "Skipped Exon (SE)")) +
    guides(fill = guide_legend(title = "Splicing Event")) +
    theme(legend.title = element_text(size = 16, face = "bold"),
          legend.text = element_text(size = 14))
  
  return(p.pie)
}
my_colors <- c("#F4F7AE", "#E6BDF4", "#FC9272", "#8E90D8", "#B9F1D6")

summary_all <- data.frame(read.csv("results/all_samples_variable_ind/output/summary.txt", header = TRUE, sep = "\t"))

# Plotting summaries
p.pieJC.tot <- nice_pie(summary_all, selected_cols = c("EventType", "TotalEventsJC"), colors = my_colors) +
  theme(legend.position = "none")
p.pieJC.sig <- nice_pie(summary_all, selected_cols = c("EventType", "SignificantEventsJC"), colors = my_colors)

p.pieJC <- p.pieJC.tot + p.pieJC.sig

p.pieJCEC.tot <- nice_pie(summary_all, selected_cols = c("EventType", "TotalEventsJCEC"), colors = my_colors) +
  theme(legend.position = "none")
p.pieJCEC.sig <- nice_pie(summary_all, selected_cols = c("EventType", "SignificantEventsJCEC"), colors = my_colors)

p.pieJCEC <- p.pieJCEC.tot + p.pieJCEC.sig
```

# 5. Volcano plots
```{r}
cutoff_alpha <- 0.05
cutoff_psi <- 0.1

# Plotting function
volcano_plot <- function(results, x_max = 1, y_max = 6, cutoff_signif = 0.05, cutoff_change = 0.1,
                         signif_var, change_var, label_var, title, colors = c("red", "grey70", "blue"))
{
  # Fixed variables
  sig_nice <- cutoff_signif/5
  chg_nice <- cutoff_change*4
  #chg_nice <- cutoff_change*2
  
  # Manage data
  d.volcano <- data.frame(results)
  d.volcano <- d.volcano[!is.na(d.volcano[, signif_var]) & d.volcano[, signif_var] < 1, ]
  d.volcano[d.volcano[, signif_var] < 10**-y_max, signif_var] <- 10**-y_max
  
  d.volcano$colors <- rep("other", nrow(d.volcano))
  d.volcano$colors[(d.volcano[, change_var] >= cutoff_change & d.volcano[, signif_var] < cutoff_signif)] <- "over"
  d.volcano$colors[(d.volcano[, change_var] <= -cutoff_change & d.volcano[, signif_var] < cutoff_signif)] <- "under"
  
  d.volcano$shapes <- rep("nohits", nrow(d.volcano))
  d.volcano$shapes[d.volcano[, signif_var] <= 10**-y_max] <- "hits"
  
  # Condition for labeling
  cond <- d.volcano[, signif_var] < sig_nice & abs(d.volcano[, change_var]) > chg_nice
  
  # Plot
  p.volcano <- ggplot() + theme_bw() +
    scale_y_continuous(expand = c(0.03, 0),
                       limits = c(0, y_max),
                       breaks = seq(0, y_max, by = 2),
                       minor_breaks = seq(0, y_max, by = 1)) +
    scale_x_continuous(expand = c(0.03, 0.03),
                       limits = c(-x_max, x_max),
                       breaks = seq(-x_max, x_max, by = x_max*0.5),
                       minor_breaks = seq(-x_max, x_max, by = x_max*0.1)) +
    
    # Vertical lines and labels:
    geom_vline(xintercept = c(-cutoff_change, cutoff_change), color = "grey20", alpha = 0.8, linetype = 2, linewidth = 1.2) +
    
    # Horizontal line and label:
    geom_hline(yintercept = -log10(cutoff_signif), color = "grey20", alpha = 0.8, linetype = 2, linewidth = 1.2) +
    geom_text(aes(x = -0.8, y = -log10(cutoff_signif) + 0.15, label = paste("q =", cutoff_signif)), color = "grey20", size = 6.2) +
    
    # Datapoints, in different colors:
    geom_point(data = d.volcano,
               aes(x = .data[[change_var]], y = -log10(.data[[signif_var]]), fill = colors, shape = shapes),
               size = 5.5, color = "gray10", alpha = 0.4, show.legend = TRUE) +
    
    # Color settings
    scale_fill_manual(name = "Splicing change",
                      values = c("over" = colors[1], "other" = colors[2], "under" = colors[3]),
                      breaks = c("over", "other", "under"),
                      labels = c("High in Tumor", "No difference", "High in Normal"),
                      guide = guide_legend(override.aes = aes(shape = 21, size = 3, alpha = 1))) +
    
    # Change outliers shapes
    scale_shape_manual(name = "Shape",
                       values = c("hits" = 24, "nohits" = 21),
                       guide = guide_legend(override.aes = list(fill = "white", size = 3, alpha = 1)),
                       labels = c(bquote(FDR <= 10^-.(y_max)), bquote(FDR > 10^-.(y_max)))) +
    
    # Labels for detectable genes
    geom_label_repel(data = d.volcano[cond, ],
                     aes(x = .data[[change_var]], y = -log10(.data[[signif_var]]), label = .data[[label_var]]),
                     inherit.aes = FALSE, parse = FALSE, max.iter = 5000, color = "black", size = 5,
                     segment.alpha = 1.5, box.padding = unit(0.8, "lines"), min.segment.length = unit(0.01, "lines")) +
    
    # Axis labels:
    labs(x = expression("Δ(PSI)"),
         y = expression("-log"[10]*"(FDR)"),
         title = title) +
    
    # Labels size
    theme(plot.title = element_text(size = 26, hjust = 0.5),
          axis.text.x = element_text(size = 22),
          axis.text.y = element_text(size = 22),
          axis.title.x = element_text(size = 26),
          axis.title.y = element_text(size = 26),
          legend.text = element_text(size = 18),
          legend.title = element_text(size= 20, face = "bold"))
  
  return(p.volcano)
}

# JC model volcanos
pv.SE.JC <- volcano_plot(results = res.SE.JC, signif_var = "FDR", change_var = "IncLevelDifference", label_var = "geneSymbol", title = "Skipped Exon")

pv.MXE.JC <- volcano_plot(results = res.MXE.JC, signif_var = "FDR", change_var = "IncLevelDifference", label_var = "geneSymbol", title = "Mutually Exclusive Exons")

pv.RI.JC <- volcano_plot(results = res.RI.JC, signif_var = "FDR", change_var = "IncLevelDifference", label_var = "geneSymbol", title = "Retained Intron")

pv.A3SS.JC <- volcano_plot(results = res.A3SS.JC, signif_var = "FDR", change_var = "IncLevelDifference", label_var = "geneSymbol", title = "Alt. 3' Splice Sites")

pv.A5SS.JC <- volcano_plot(results = res.A5SS.JC, signif_var = "FDR", change_var = "IncLevelDifference", label_var = "geneSymbol", title = "Alt. 5' Splice Sites")

# JCEC model volcanos
pv.SE.JCEC <- volcano_plot(results = res.SE.JCEC, signif_var = "FDR", change_var = "IncLevelDifference", label_var = "geneSymbol", title = "Skipped Exon")

pv.MXE.JCEC <- volcano_plot(results = res.MXE.JCEC, signif_var = "FDR", change_var = "IncLevelDifference", label_var = "geneSymbol", title = "Mutually Exclusive Exons")

pv.RI.JCEC <- volcano_plot(results = res.RI.JCEC, signif_var = "FDR", change_var = "IncLevelDifference", label_var = "geneSymbol", title = "Retained Intron")

pv.A3SS.JCEC <- volcano_plot(results = res.A3SS.JCEC, signif_var = "FDR", change_var = "IncLevelDifference", label_var = "geneSymbol", title = "Alt. 3' Splice Sites")

pv.A5SS.JCEC <- volcano_plot(results = res.A5SS.JCEC, signif_var = "FDR", change_var = "IncLevelDifference", label_var = "geneSymbol", title = "Alt. 5' Splice Sites")
```

# 6. Heatmaps by splicing event
```{r}
colnames <- data.frame(label = c("51", "57", "59", "63", "148",
                                 "51P", "51-M1", "51-M2", "51-M3", "57M", "59P", "59M",
                                 "62M", "63R", "88-M1", "88-M2", "117M", "123M", "148-P1", "148-P2"),
                       group = c(rep("Normal", 5), rep("Tumor", 15)))

# Colors
ann_colors <- list(Group = c(Normal = "#bbffff", Tumor = "#ffb6c1"))

# Annotation
group_labels <- c(rep("Normal", 5), rep("Tumor", 15))
anot <- data.frame(Group = factor(group_labels))

# Extract psi values
split_psi <- function(df, gene_col = "geneSymbol", psi_cols = c("IncLevel1", "IncLevel2"), num = 10) {
  initial_df <- df[order(df$FDR), ]
  initial_df <- initial_df[1:num, ]
  psi_split <- lapply(initial_df[, psi_cols], function(x) t(sapply(strsplit(x, ","), as.numeric)))
  psi_mat <- do.call(cbind, psi_split)
  rownames(psi_mat) <- initial_df[[gene_col]]
  return(as.data.frame(psi_mat))
}

# Test data frame
mx.SE.JC <- split_psi(res.SE.JC.sig)
mx.SE.JC <- mx.SE.JC[rowSums(is.na(mx.SE.JC)) != ncol(mx.SE.JC), ]
mx.SE.JCEC <- split_psi(res.SE.JCEC.sig)
mx.SE.JCEC <- mx.SE.JCEC[rowSums(is.na(mx.SE.JCEC)) != ncol(mx.SE.JCEC), ]

mx.MXE.JC <- split_psi(res.MXE.JC.sig)
mx.MXE.JC <- mx.MXE.JC[rowSums(is.na(mx.MXE.JC)) != ncol(mx.MXE.JC), ]
mx.MXE.JCEC <- split_psi(res.MXE.JCEC.sig)
mx.MXE.JCEC <- mx.MXE.JCEC[rowSums(is.na(mx.MXE.JCEC)) != ncol(mx.MXE.JCEC), ]

mx.RI.JC <- split_psi(res.RI.JC.sig)
mx.RI.JC <- mx.RI.JC[rowSums(is.na(mx.RI.JC)) != ncol(mx.RI.JC), ]
mx.RI.JCEC <- split_psi(res.RI.JCEC.sig)
mx.RI.JCEC <- mx.RI.JCEC[rowSums(is.na(mx.RI.JCEC)) != ncol(mx.RI.JCEC), ]

mx.A3SS.JC <- split_psi(res.A3SS.JC.sig)
mx.A3SS.JC <- mx.A3SS.JC[rowSums(is.na(mx.A3SS.JC)) != ncol(mx.A3SS.JC), ]
mx.A3SS.JCEC <- split_psi(res.A3SS.JCEC.sig)
mx.A3SS.JCEC <- mx.A3SS.JCEC[rowSums(is.na(mx.A3SS.JCEC)) != ncol(mx.A3SS.JCEC), ]

mx.A5SS.JC <- split_psi(res.A5SS.JC.sig)
mx.A5SS.JC <- mx.A5SS.JC[rowSums(is.na(mx.A5SS.JC)) != ncol(mx.A5SS.JC), ]
mx.A5SS.JCEC <- split_psi(res.A5SS.JCEC.sig)
mx.A5SS.JCEC <- mx.A5SS.JCEC[rowSums(is.na(mx.A5SS.JCEC)) != ncol(mx.A5SS.JCEC), ]

mx.JC.merged <- rbind(mx.SE.JC, mx.MXE.JC, mx.RI.JC, mx.A3SS.JC, mx.A5SS.JC)
mx.JCEC.merged <- rbind(mx.SE.JCEC, mx.MXE.JCEC, mx.RI.JCEC, mx.A3SS.JCEC, mx.A5SS.JCEC)
colnames(mx.JC.merged) <- colnames$label
colnames(mx.JCEC.merged) <- colnames$label

# Plot heatmap
plot_heatmap <- function(df, annotation = anot, annotation_cols = ann_colors) {
  psi_scaled <- t(scale(t(df)))
  rownames(annotation) <- colnames(psi_scaled)
  
  p.heat <- pheatmap(psi_scaled,
                     cluster_cols = FALSE,
                     cluster_rows = FALSE,
                     show_rownames = TRUE, 
                     scale = "none",
                     gaps_row = c(10, 20, 30, 40),
                     gaps_col = c(5),
                     annotation_col = annotation,
                     annotation_colors = annotation_cols,
                     fontsize_row = 8,
                     show_colnames = TRUE)
  
  return(p.heat)
}

ph.JC.merged <- plot_heatmap(mx.JC.merged)
ph.JCEC.merged <- plot_heatmap(mx.JCEC.merged)
```

# 7. [IN PROGRESS] Barplots per gene with splicing events
```{r}
count_shared_events <- function(df, psi_col, group) {
  # Split and convert PSI strings to numeric vectors
  psi_matrix <- t(sapply(strsplit(df[[psi_col]], ","), as.numeric))
  psi_df <- as.data.frame(psi_matrix)
  
  # Assign consistent sample names
  colnames(psi_df) <- paste(group, seq_len(ncol(psi_df)), sep = "_")
  
  # Bind gene symbol and event ID (optional)
  psi_df$geneSymbol <- df$geneSymbol
  psi_df$eventID <- seq_len(nrow(df))  # to track unique splicing events if needed
  
  # Count number of non-NA & non-zero PSI values per event (row)
  count_df <- psi_df[, 1:(ncol(psi_df) - 2)]  # exclude geneSymbol and eventID
  non_zero_counts <- rowSums(!is.na(count_df) & count_df > 0)
  
  # Create frequency table
  frequency_table <- table(non_zero_counts)
  
  return(list(table = frequency_table, matrix = psi_df))
}


```

# 8. Map peptides to splicing variants
## Call libraries
```{r}
library(GenomicFeatures)
library(BSgenome.Hsapiens.UCSC.hg38)
library(Biostrings)
library(GenomicRanges)
```


## Read reference
```{r}
tx_tab <- data.frame(read.xlsx("../../../Documents/Articles/CCC/tx2gene_GRCh38v103.xlsx"))
gtf_file <- "GRCh38.103.gtf"

txdb <- txdbmaker::makeTxDbFromGFF(file = gtf_file, format = "gtf",
                                   organism = "Homo sapiens", dataSource = "Ensembl_103")
```

## Build isoform cDNA
```{r}
buildSkippingIsoformCDNA <- function(se_row, tx_table = tx_tab, edb = EnsDb.Hsapiens.v103,
                                     bsgenome = BSgenome.Hsapiens.UCSC.hg38)
{
  # Extract rMATS coordinates
  chr <- as.character(se_row$chr)
  se_start <- se_row$exonStart_0base + 1
  se_end <- se_row$exonEnd
  
  gene_sym <- as.character(se_row$geneSymbol)
  
  # Filter ensemblDB
  transcripts_for_gene <- tx_table[tx_table$symbol == gene_sym, "transcriptID"]
  transcripts_for_gene <- unique(as.character(transcripts_for_gene))
  
  if (length(transcripts_for_gene) == 0) {
    stop("No transcriptID for symbol ", gene_sym)
  }
  
  # Obtain exon GRangesList by transcriptIDs
  exon_list_all <- exonsBy(edb, by = "tx", filter = TxIdFilter(transcripts_for_gene))
  
  skip_cdna_list <- DNAStringSet()
  
  # Process exons
  for (tx_id in names(exon_list_all)) {
    exons_of_tx <- exon_list_all[[tx_id]]
    
    # Verify if any exon exactly matches the SE
    hits <- which(
      as.character(seqnames(exons_of_tx)) == chr &
      start(exons_of_tx) == se_start &
      end(exons_of_tx) == se_end
    )
    if (length(hits) == 0) {
      next
    }
    
    # Remove identified exon
    exons_skip <- exons_of_tx[-hits]
    exons_skip <- sort(exons_skip)
    
    # Extract DNA from concatenated exons
    chr_seq <- bsgenome[[as.character(seqnames(exons_skip)[1])]]
    ranges_list <- GRanges(
      seqnames = seqnames(exons_skip),
      ranges = IRanges(start = start(exons_skip),
                       end = end(exons_skip)),
      strand = strand(exons_skip)
    )
    cds_dna <- getSeq(bsgenome, ranges_list)
    skip_cdna <- do.call(xscat, as.list(cds_dna))
    
    # Managing reverse compiment (- strand)
    tx_strand <- as.character(strand(exons_skip)[1])
    if (tx_strand == "-") {
      skip_cdna <- reverseComplement(skip_cdna)
    }
    
    # Save transcript ID
    skip_cdna_list <- append(skip_cdna_list, skip_cdna)
    names(skip_cdna_list)[length(skip_cdna_list)] <- tx_id
  }
  
  if (length(skip_cdna_list) == 0) {
    warning("No transcript from ", gene_sym, " contained coordinates: ", se_start, "-", se_end, ".")
  }
  return(skip_cdna_list)
}
```

## Function to tile peptides
```{r}
i <- 23083
se_row <- res.SE.JC.sig[i, ]

# Generate skipped isoform
skip_cdna_set <- buildSkippingIsoformCDNA(se_row)

# Translate isoform
translateSkippingIsoform <- function(skip_cdna) {
  rna_seq <- RNAString(skip_cdna)
  aa_seq  <- translate(rna_seq, if.fuzzy.codon="solve")
  return(aa_seq)
}

# Get junction coordinates
findJunctionAADist <- function(skipped_exons_gr, tx_strand, se_row) {
  exon_widths <- width(skipped_exons_gr)
  
  de_start <- se_row$downstreamES
  de_idx <- which(start(skipped_exons_gr) <= de_start & end(skipped_exons_gr) >= de_start)
  if (length(de_idx) != 1) {
    stop("Downstream exon not identified for: ", se_row$geneSymbol)
  }
  
  if (tx_strand == "+") {
    offset_within_de <- de_start - start(skipped_exons_gr[de_idx]) + 1
  } else {
    offset_within_de <- end(skipped_exons_gr[de_idx]) - de_start + 1
  }
  
  if (de_idx > 1) {
    total_bases_before <- sum(exon_widths[1:(de_idx-1)]) + offset_within_de - 1
  } else {
    total_bases_before <- offset_within_de - 1
  }
  
  aa_pos <- floor(total_bases_before / 3) + 1
  return(list(n_bases = total_bases_before, aa_index = aa_pos))
}

# Tile neoepitopes
tileNeoepitopes <- function(aa_seq, aa_index, min_len = 8, max_len = 12) {
  peptides <- data.frame(length = integer(),
                         start_AA = integer(),
                         peptideSeq = character(),
                         stringsAsFactors = FALSE)
  seq_length <- length(aa_seq)
  
  for (L in min_len:max_len) {
    possible_starts <- (aa_index - (L - 1)):aa_index
    for (s in possible_starts) {
      e <- s + L - 1
      if (s >= 1 && e <= seq_length) {
        pep <- as.character(subseq(aa_seq, start = s, end = e))
        peptides <- rbind(peptides, data.frame(length = L, start_AA = s,
                                               peptideSeq = pep, stringsAsFactors = FALSE))
      }
    }
  }
  peptides <- unique(peptides)
  return(peptides)
}

```

## Obtain final data frame
```{r}
all_results <- list()

for (tx_id in names(skip_cdna_set)) {
  this_cdna <- skip_cdna_set[[tx_id]]
  aa_seq <- translateSkippingIsoform(this_cdna)
  
  # Retrieve original exon
  exons_for_tx <- exonsBy(EnsDb.Hsapiens.v103, by = "tx",
                          filter = TxIdFilter(tx_id))[[1]]
  
  # Filter SE
  hits <- which(start(exons_for_tx) == (se_row$exonStart_0base + 1) &
                  end(exons_for_tx) == se_row$exonEnd)
  exons_skip <- exons_for_tx[-hits]
  exons_skip <- sort(exons_skip)
  
  chr_strand <- as.character(strand(exons_skip)[1])
  junction_info <- findJunctionAADist(exons_skip, chr_strand, se_row)
  aa_index <- junction_info$aa_index
  
  neopep_df <- tileNeoepitopes(aa_seq, aa_index, min_len = 8, max_len = 12)
  neopep_df$transcript <- tx_id
  neopep_df$gene <- se_row$geneSymbol
  neopep_df$aa_index <- aa_index
  neopep_df$event_id <- se_row$ID
  
  all_results[[tx_id]] <- neopep_df
}

final_neoepitopes <- do.call(rbind, all_results)
rownames(final_neoepitopes) <- NULL
head(final_neoepitopes)

```

